---
title: "District 7 Campaign Contribution Analysis"
author: "Larkin O'Hern"
date: "January 25, 2017"
output: html_document
---
```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}

###load packages
library(tidyverse)
library(lubridate)
library(forcats)
library(knitr)
library(stringr)

###read in the data 

raw_data<-"dist7_contribution_report_jan2017.csv" ###insert file name here

campaign_finance<-as_tibble(read.csv(raw_data))

```

```{r, echo=FALSE, include=FALSE, message=F, warning=F}

campaign_finance1<-na.omit(campaign_finance) ###na.omit() to drop empty rows created by the two line adresses


campaign_finance2<-campaign_finance1 %>% ###this will be our primary tibble going forward
        select(-Report.Id.,-FilerName, -Report.Type.Code.,-Count.) %>% #drop vars that don't add value
        rename(contributor=Name., #fix names
               contact_type=Contact.Type.,
               report_type=Report.Type.,
               amount=Amount.,
               contrib_type=strVal,
               candidate=Cand.Committee.,
               election_date=Election.Date.,
               transaction_id=Id,
               report_id=ReportId,
               transaction_date=TransDate.,
               report_date=CreatedDt.) %>% 
        separate(transaction_date, into=c("transaction_date","time"), sep=" ") %>% #get dates as dates
        select(-time) %>% 
        mutate(transaction_date=mdy(transaction_date)) %>% 
        separate(report_date, into=c("report_date","time"), sep=" ") %>% 
        select(-time) %>% 
        mutate(report_date=mdy(report_date),
               report_id=factor(report_id),
               transaction_id=factor(transaction_id)
               ) %>% 
        separate(contributor, into=c("last_name","first_name"), sep=",") %>%  #first name na means PAC
        mutate(contact_type=ifelse(is.na(first_name),"PAC","INDV"))#create a var to id PACs and individuals

```

```{r echo=FALSE, warning=FALSE, message=F}
#this is a "mode" subsitute function 
most_common<-function(x){
        ux<-unique(x)
        ux[which.max(tabulate(match(x,ux)))]
}

#this section fixes the problem created when adresses were read in as two new rows in the Name. column
campaign_finance1a<- campaign_finance%>% 
        separate(Name., into=c("last_name","first_name"), sep=",")
        
        
adress<-str_subset(campaign_finance1a$last_name,"\\d")        

zipcode<-(str_subset(campaign_finance1a$first_name,"\\d"))

zipcode1<-as_tibble(trimws(zipcode))

zipcode1<-separate(zipcode1,value, into=c("state", "zipcode"), sep ="  ")  #create new state var. I'm suprised at the number of out of state contributions for a mayoral race.

#zipcode1<-na.omit(zipcode1) ###5 adresses with odd formatting dropped Bldg # ect.

campaign_finance2a<-as_tibble(cbind(adress,zipcode1, campaign_finance2)) #this will be the primary tibble for geo data.
        
unique_adresses<-length(unique(as.character(campaign_finance2a$adress))) #thi will id unique donor families

unique_indv_contributors<-distinct(campaign_finance2a, adress, first_name,last_name, contact_type, candidate) %>% 
        filter(contact_type=="INDV") %>% 
        nrow()

        
contributing_households<-distinct(campaign_finance2a, adress, contact_type) %>% 
        filter(contact_type=="INDV") %>% 
        nrow()

multi_givers<-select(campaign_finance2a, adress, last_name, first_name, contact_type, amount, candidate) %>% 
        filter(contact_type=="INDV") %>% 
        group_by(first_name, last_name, adress) %>% 
        summarize(n=n(),
                  total=sum(amount),
                  mode=most_common(amount)) %>% 
        filter(n>=2) %>% 
        arrange(-total)

over_maxers<-multi_givers %>% 
        filter(total>1000) %>% 
        left_join(campaign_finance2a) ###it appears Victoria Rico has cotributed above the $1000 limit 
                                        ###to Mayor Ivy
        
give_to_two_canidates<-select(campaign_finance2a, adress, last_name, first_name, contact_type, amount, candidate) %>% 
        group_by(first_name, last_name, adress) %>% 
        summarize(n=n(),
                  total=sum(amount),
                  mode=most_common(amount)) %>% 
        filter(n>=2) %>% 
        left_join(campaign_finance2a) %>%   
        select(adress, last_name, first_name, contact_type, amount, candidate)
        




candidate_summary<-campaign_finance2 %>% 
        group_by(candidate) %>% 
        summarise(total=sum(amount),
                  n=n(),
                  mean=mean(amount),
                  median=median(amount),
                  mode=most_common(amount),
                  max=max(amount),
                  min=min(amount)
                  ) %>% 
        arrange(-total)
```
### The Data

```{r echo=FALSE}
num_candidates=(length(levels(campaign_finance2$candidate)))-1 #ugly way to eliminate the 0 candidate need a dplyr drop()

cand1<-levels(campaign_finance2$candidate)[2]
cand2<-levels(campaign_finance2$candidate)[3]
cand3<-levels(campaign_finance2$candidate)[4]
cand4<-levels(campaign_finance2$candidate)[5]
cand5<-levels(campaign_finance2$candidate)[6]

contrib_type_table<-table(campaign_finance2$contact_type)

first_contribution<-head(sort(campaign_finance2$transaction_date),1)[1]
last_contribution<-tail(sort(campaign_finance2$transaction_date),1)[1]
num_contributions<-length(campaign_finance2$candidate) ###neeed to eliminate duplicates
num_indv_contributors<-contrib_type_table[1]
num_pacs<-contrib_type_table[2]
total_contributions<-sum(campaign_finance2$amount)

options(scipen = 5, digits = 3  )

```

This data for this report is available here: and is up to date as of  **`r campaign_finance2$report_date[1]`**.  At that time there were **`r num_candidates`** candidates: **`r cand1`**, 
**`r cand2`**, **`r cand3`**, **`r cand4`** and **`r cand5`**.  The report included data on **`r num_contributions`** contributions from **`r num_indv_contributors`** individual contributors and **`r num_pacs`** Political Action Committees worth **$`r total_contributions`** recorded between **`r first_contribution`** and **`r last_contribution`**.

### PACs vs. Individuals (Families?)
```{r echo=FALSE}
pacs_v_indv<-campaign_finance2 %>% 
        group_by(contact_type) %>% 
        summarize(total=sum(amount),
                  mean=mean(amount),
                  median=median(amount),
                  mode=most_common(amount),
                  max=max(amount),
                  min=min(amount)
                  )

kable(pacs_v_indv, digits=2, format.args=list(big.mark=","))

```


Individual donors outspent PACs **$ `r pacs_v_indv[1,2]`** to **$`r pacs_v_indv[2,2]`** but the average  PAC gave significantly more- **$`r pacs_v_indv[2,3]`** to **$`r pacs_v_indv[1,3]`** (mean) and **$`r pacs_v_indv[2,3]`** to **$`r pacs_v_indv[1,3]`** (median).  The most common individual contribution was **$`r pacs_v_indv[1,5]`** while the most common PAC contribution was **$`r pacs_v_indv[2,5]`**. See Appendix A for a list of all PAC contributions. 

Thus far **`r unique_indv_contributors`** individuals from **`r contributing_households`** households have made contributions. **`r nrow(multi_givers)`** individuals have made more than one contribution.  
```{r, echo=FALSE}

ggplot(campaign_finance2, aes(x=contact_type, y=amount))+
        geom_violin()+
        geom_boxplot(width=.1, fill="black")+
        stat_summary(fun.y=mean, geom = "point", fill="white", size=2.5, shape=21)+
        xlab("Type of Contributor")+ylab("Contribution Amount ($)")+ggtitle("Distribution of  Contribution Amounts by Contributor Type")

```

##The Candidates

```{r, echo=F, include=T}


ggplot(campaign_finance2, aes(x=candidate, y=amount))+
        geom_violin()+
        #geom_boxplot(width=.01, fill="black")+
        stat_summary(fun.y=mean, geom = "point", fill="black", size=2, shape=21)+
        xlab("Candidate")+ylab("Contribution Amount ($)")+ggtitle("Distribution of Contribution Amounts by Candidate")

kable(candidate_summary, digits=2, format.args=list(big.mark=","))


contributions_over_time<-campaign_finance2 %>% 
        mutate(candidate=fct_reorder(candidate, -amount)) %>% 
        group_by(candidate, transaction_date) %>% 
        summarize(daily_total=sum(amount)) %>% 
        mutate(running_total=cumsum(daily_total))

ggplot(contributions_over_time, aes(x=transaction_date, y=running_total, color=candidate))+ geom_line()+xlab("Date")+ylab("Total Contributions ($)")+ggtitle("Total Contributions From 01 July-31 DEC 2016")+theme(legend.position="bottom")+labs(color="")

maxers<-campaign_finance2 %>%
        filter(amount==1000) %>% 
        group_by(candidate, contact_type) %>%
        summarize(maxers=n())

pac_giving<-campaign_finance2 %>% 
        group_by(candidate, contact_type) %>% 
        summarise(pac_contributors=n()) %>% 
        filter(contact_type=="PAC") %>% 
        select(-contact_type) %>% 
        arrange(-pac_contributors)

giving_by_zipcode<-campaign_finance2a %>% 
        group_by(candidate, zipcode) %>% 
        summarise(total=sum(amount),
                  mean=mean(amount),
                  n=n()) %>% 
        arrange(-total) %>% 
        left_join(candidate_summary, by="candidate") %>% 
        select(-median, -mode,-min,-max) %>%
        mutate(percent=(total.x/total.y)*100,
               rank=min_rank(desc(total.x))) %>% 
        filter(rank<=10)
        
ggplot(giving_by_zipcode)+
        geom_bar(aes(x=reorder(zipcode,total.x),y=total.x, fill=fct_reorder(candidate, total.y)), stat="identity")+
        facet_wrap(~candidate)+
        coord_flip()+
        ylab("Total Contributions ($)")+xlab("Zip Code")+ggtitle("Contributions by Zip Code")+
        scale_fill_discrete(guide=F)


family_giving<-campaign_finance2a %>% 
        group_by(adress, state, zipcode, last_name, candidate) %>% 
        summarize(n=n(),
                  total_giving=sum(amount)) %>% 
        arrange(-total_giving) %>% 
        filter(n>=2)


out_of_state_giving_by_state<-campaign_finance2a %>%
        filter(state!="TX") %>% 
        group_by(candidate, state) %>% 
        summarize(total=sum(amount),
                  mean=mean(amount),
                  n=n()) %>% 
        arrange(-total)

out_of_state_giving_totals<-campaign_finance2a %>%
        filter(state!="TX") %>% 
        group_by(candidate, contact_type) %>% 
        summarize(total=sum(amount),
                  mean=mean(amount),
                  n=n()) %>% 
        arrange(-total) %>% 
        left_join(candidate_summary, by="candidate") %>% 
        select(candidate:total.y) %>% 
        rename(total=total.x, mean=mean.x, n=n.x,grand_total=total.y) %>% 
        mutate(percent_grand_total=(total/grand_total)*100)
```


*Total Out of State Donations:*
`r kable(out_of_state_giving_totals, digits=2, format.args=list(big.mark=","))`

*Out of State Donations By State:*
`r kable(out_of_state_giving_by_state, digits=2, format.args=list(big.mark=","))`


*Max ($1000) Donators*
`r kable(maxers,digits=2, format.args=list(big.mark=","))`

*PAC Giving*
`r kable(pac_giving,digits=2, format.args=list(big.mark=","))`

```

### Contributions by Type (Monetary vs. In-Kind)

```{r, echo=FALSE}

###lets look at contribution types 

contrib_type_by_candidate<-campaign_finance2 %>% 
        group_by(candidate,contact_type, contrib_type) %>% 
        summarize(total=sum(amount),
                  mean=mean(amount),
                  n=n()) %>% 
        arrange(-total)

ggplot(contrib_type_by_candidate, aes(x=candidate, y=total, fill=contrib_type))+
        geom_bar(stat="identity",position = "dodge")+guides(fill=FALSE)

kable(contrib_type_by_candidate, digits=2, format.args=list(big.mark=",") )

```

##Appendix A-PAC Contributions 
```{r, echo=FALSE, include=TRUE}

pac_contributions<-campaign_finance2 %>% 
        filter(contact_type=="PAC") %>% 
        select(last_name, amount, candidate) %>% 
        arrange(candidate, -amount)



kable(pac_contributions, digits=2, format.args=list(big.mark=","))
        
```